---
title: "Re-mining pipeline"
output: html_document
---

```{r}
screen <- "twoPlates"
channel <- "nucleus-dapi"
```

> **SCREEN:    `r screen`**

> **CHANNEL:   `r channel`**

> **PLATE:     `r plate`**

> **REPLICATE: `r replicate`**

------------------------------------------------------------------------------------------------------------------------------

------------------------------------------------------------------------------------------------------------------------------

### **Step 0: Set up the environment management**

------------------------------------------------------------------------------------------------------------------------------

- Screen settings.
```{r, echo=F}
if (screen == "mockScreen")
{
  INDEX_START_FEATURE_DATA <- 15
  knitr::opts_chunk$set(fig.width=10, fig.height=8)
}
if (screen == "secretionPilot")
{
  INDEX_START_FEATURE_DATA <- 16
  knitr::opts_chunk$set(fig.width=14, fig.height=12)
}
if (screen == "twoPlates")
{
  INDEX_START_FEATURE_DATA <- 14
  knitr::opts_chunk$set(fig.width=16, fig.height=14)
}
```
- Load paths.
```{r, echo=F}
source("paths.R")
```
- Load functions.
```{r, echo=F}
# source("libs/mylib_images.R")
# source("libs/mylib_matrixAlgebra.R")
# source("libs/mylib_measures.R")
# source("libs/mylib_plots.R")
source("libs/mylib_stats.R")
# source("libs/mylib_utils.R")
```
- Load packages.
```{r, message=F, warning=F, echo=T}
#library(ggdendro)
#library(ggplot2)
#library(gridExtra)
#library(igraph)
#library(pheatmap)
#library(rARPACK)
#library(e1071)
#library(kernlab)
#library(tsne)
#library(Rtsne)
#library(caret)
library(mclust)
```
- Define constants.
```{r, echo=F}
LABEL_CONTROL_CLASS       <- "notarget"
LABEL_NO_PHENOTYPE        <- "nophenotype"
LABEL_CONTROL_TAG         <- "CTRL"
LABEL_NO_PHENOTYPE_TAG    <- "NO_PH"
LINKAGE_METHOD_FOR_HCLUST <- "average"
PROB_IMAGE_BELONG_CONTROL <- 0.7
```

------------------------------------------------------------------------------------------------------------------------------

### **Step 1: Feature extraction from images (WND-CHARM)**

------------------------------------------------------------------------------------------------------------------------------

- Data file must have the following column structure:
    - First part: metadata, where column order is not relevant, except for *class* that must be at the end.
    - Second part: undetermined number of features from WND-CHARM in any order.

```{r, echo=F}
dataPath <- file.path(pathTo_data, paste0(screen,"_db_",channel,".txt"))
data <- read.csv(dataPath, header=T, sep="\t", check.names=F)
```

- Select plate and replicate:
```{r}
data <- subset(data, rep==replicate & Plate==plate)
if (nrow(data) == 0)
{
    cat(paste("There are no images for replicate", replicate, "and plate", plate))
    knit_exit()
}
```

- Standarize naming: TEST CLASS WITH NO PHENOTYPE.
```{r, echo=F}
# Elements with no phenotype (from test images) should have as label "NO_PH" and IC=0
if (length(data[data$specificCMPO==LABEL_NO_PHENOTYPE,]$IC) > 0)
{
  data[data$specificCMPO==LABEL_NO_PHENOTYPE,]$IC <- 0
  levels(data$tag) <- c(levels(data$tag), LABEL_NO_PHENOTYPE_TAG)
  data[data$specificCMPO==LABEL_NO_PHENOTYPE,]$tag <- LABEL_NO_PHENOTYPE_TAG
}
```

- Standarize naming: CONTROL CLASS.
```{r, echo=F}
# Those with empty gene are no targeted genes
if (length(data[data$symbol=="",]$symbol) > 0)
{
  data[data$symbol=="",]$symbol <- LABEL_CONTROL_CLASS
}
if (nrow(data[data$symbol=="",]) > 0)
{
  data[data$symbol==LABEL_CONTROL_CLASS,]$specificName <- LABEL_NO_PHENOTYPE
  data[data$symbol==LABEL_CONTROL_CLASS,]$specificCMPO <- LABEL_CONTROL_CLASS
  data[data$symbol==LABEL_CONTROL_CLASS,]$EnsemblID    <- LABEL_CONTROL_CLASS
  data[data$symbol==LABEL_CONTROL_CLASS,]$IC           <- 0
  data[data$symbol==LABEL_CONTROL_CLASS,]$tag          <- LABEL_CONTROL_TAG
  data[data$symbol==LABEL_CONTROL_CLASS,]$class        <- LABEL_CONTROL_CLASS
}
```

- Build an unique ID to identify images (*data$id*): **gene_siRNA_well**.
```{r, echo=F}
data$id <- paste(data$symbol, data$siRNA, data$Well, sep="_") 
data <- data[c(ncol(data), 1:(ncol(data)-1))]
INDEX_START_FEATURE_DATA <- INDEX_START_FEATURE_DATA + 1

colsMetadata <- colnames(data)[1:(INDEX_START_FEATURE_DATA-1)]
data.metadata <- data[c(colsMetadata[colsMetadata!="class"],colsMetadata[colsMetadata=="class"])]
data.features <- data[c(INDEX_START_FEATURE_DATA:ncol(data))]
data <- cbind(data.metadata, data.features)
rm(data.metadata, data.features)
```

- Number of images for each siRNA:
```{r}
sirna_freq <- as.data.frame(table(data$siRNA))
colnames(sirna_freq) <- c("siRNA", "#images")
```

- Identify positive controls:
```{r}
positiveControls <- subset(sirna_freq, `#images`>1)$siRNA
# Remove negative control
positiveControls <- positiveControls[positiveControls!="103860"]
# Mark siRNA and gene names
data[data$siRNA %in% positiveControls,]$id <- paste0(data[data$siRNA %in% positiveControls,]$id,"**********")
```

- Set colors for plots.
```{r, echo=F}
colorCodes <- unique(data[c("id","tag")])
colnames(colorCodes) <- c("label","tag")
```

- Structure pointers to adress data:
```{r, echo=F}
dataColNames <- colnames(data)
indexID    <- match("id", dataColNames)
indexClass <- match("class", dataColNames) 
indexRangeFeature <- INDEX_START_FEATURE_DATA:ncol(data)
```

```{r, echo=F}
# Show first images
# knitr::kable(data[1:16, c("id","specificCMPO","tag",indexClass,indexRangeFeature)])
```

> THERE ARE **`r nrow(data)`** IMAGES.

```{r}
table(data$class)
table(data$tag)
table(data$specificCMPO)
```

------------------------------------------------------------------------------------------------------------------------------

### **Step 2: Feature selection**

------------------------------------------------------------------------------------------------------------------------------

#### **2.1. Feature initial filtering**

- WND-CHARM features:
```{r}
originalFeatures <- data[,indexRangeFeature]
```

- Remove columns with standard deviation = 0:
```{r}
features_noConstants <- stats.remove_standardDeviation0_fromDF(originalFeatures)
featureNamesToBeSelected <- names(features_noConstants)
```

- Update data and pointers.
```{r, echo=F}
metadata <- data[1:(INDEX_START_FEATURE_DATA-1)]
features <- data[featureNamesToBeSelected]

data <- data.frame(metadata, features, check.names=F)
indexRangeFeature <- c((ncol(metadata)+1):ncol(data))
```

#### **2.2. PCA for feature selection**

```{r, fig.width=20, fig.height=8}
pca <- prcomp(data[indexRangeFeature], scale.=T, center=T)
pcaEigenvalues <- pca$sdev^2
plot(pcaEigenvalues[1:length(pcaEigenvalues)], type="l", main=paste0("VARIANCES 1-",length(pcaEigenvalues)))
numberOfSelectedFeatures <- 10
pcaEigenvalues <- pcaEigenvalues[1:numberOfSelectedFeatures]
abline(v=numberOfSelectedFeatures, col="red")
```

| Stage                                           | # features                             |
| ----------------------------------------------- |:--------------------------------------:|
| After WND-CHARM feature extraction              | `r ncol(originalFeatures)`             |
| After cleaning constant columns                 | `r ncol(features_noConstants)`         | 
| After PCA                                       | `r numberOfSelectedFeatures`           |
|                                                 |                                        | 

- Selected features:
```{r}
filteredFeatures <- as.data.frame(pca$x[,1:numberOfSelectedFeatures])
metadata <- data[1:(INDEX_START_FEATURE_DATA-1)]
```

- Update data and pointer with the new PCA features:
```{r}
data <- data.frame(metadata, filteredFeatures, check.names=F)
indexRangeFeature <- INDEX_START_FEATURE_DATA:ncol(data)
```

- Save selected features.
```{r, echo=F}
cleanFeaturesFile <- file.path(pathTo_data, paste0(screen,"_db_filteredFeatures_",channel,".txt"))
write.table(data, file=cleanFeaturesFile, sep="\t", col.names=T, row.names=T, append=F, quote=F)
```


------------------------------------------------------------------------------------------------------------------------------

### **Step 3: Outlier detection and data reduction**

------------------------------------------------------------------------------------------------------------------------------

- Split data into control and test.
```{r, echo=F}
controlClass <- subset(data, data$class==LABEL_CONTROL_CLASS)
testClass    <- subset(data, data$class!=LABEL_CONTROL_CLASS)

rownames(controlClass) <- controlClass$id
rownames(testClass)    <- testClass$id

controlClassFeatures   <- controlClass[indexRangeFeature]
testClassFeatures      <- testClass[indexRangeFeature]
```

> THERE ARE **`r nrow(testClass)`** TEST IMAGES AND **`r nrow(controlClass)`** CONTROL IMAGES.

#### **3.1. Gaussian Mixture Model**

##### **3.1.1. Model for the Control Class**

```{r}
cc_gmm <- Mclust(controlClassFeatures)
numberOfClusters <- max(unique(cc_gmm$classification))
summary(cc_gmm)
```

- Estimation of the number of clusters for the control class images.
```{r}
#plot(cc_gmm, what = "classification")
plot(cc_gmm$BIC)
```

> The **best number of clusters** is **`r numberOfClusters`** (given by the peak and the decrease).

##### **3.1.2. Prediction for the test images**

```{r, fig.width=20, fig.height=10}
pred_testImages <- predict.Mclust(cc_gmm, testClassFeatures)
probs <- pred_testImages$z

if (numberOfClusters > 1)
{
  prob_test_belong_control1 <- probs[,1]
  hist(prob_test_belong_control1, cex.axis=1.5, breaks="Sturges")
  testImages_similarToControl1 <- prob_test_belong_control1[prob_test_belong_control1>PROB_IMAGE_BELONG_CONTROL]

  prob_test_belong_control2 <- probs[,2]
  hist(prob_test_belong_control2, cex.axis=1.5, breaks="Sturges")
  testImages_similarToControl2 <- prob_test_belong_control2[prob_test_belong_control2>PROB_IMAGE_BELONG_CONTROL]
}else
{
  testImages_similarToControl1 <- numeric() 
  testImages_similarToControl2 <- numeric()
}
```

| TEST IMAGES                                  | # images                                  |
| -------------------------------------------- |:-----------------------------------------:|
| Total number of test images                  | `r nrow(testClassFeatures)`               |
| Number of test images similar to control 1   | `r length(testImages_similarToControl1)`  | 
| Number of test images similar to control 2   | `r length(testImages_similarToControl2)`  |
|                                              |                                           | 
| Number of remaining test images              | `r nrow(testClassFeatures) - (length(testImages_similarToControl1) + length(testImages_similarToControl2))`                                                     | 
|                                              |                                           | 


```{r}
table(data$class)
table(data$tag)
table(data$specificCMPO)
```

#### **3.2. Data reduction**

- Reject ALL CONTROLS (**-`r nrow(testClassFeatures)`**): 

Starting with **`r nrow(data)`** images (TI + CC)...
```{r}
data <- data[data$class!=LABEL_CONTROL_CLASS,]
```
... we get **`r nrow(data)`** (TI + CC).


```{r}
table(data$tag)
table(data$specificCMPO)
```

- Reject TEST IMAGES close to artefacts (**-`r length(testImages_similarToControl1)`**) or controls (**-`r length(testImages_similarToControl2)`**):

Starting with **`r nrow(data)`** images (TI + CC)...
```{r}
imagesToReject <- as.factor(c(names(testImages_similarToControl2), names(testImages_similarToControl1)))
data <- data[!data$id %in% imagesToReject,]
if (nrow(data) == 0)
{
  cat("All images have been detected to be close to artefacts")
  knit_exit()
}
```
... we get **`r nrow(data)`**.

- Save data.
```{r, echo=F}
fileData <- file.path(pathTo_data, paste0(screen,"_db_cleanWeights_cleanArtefacts_dataReduction_",channel,".txt"))
write.table(data, file=fileData, sep="\t", col.names=F, row.names=T, append=F, quote=F)
```

> THERE ARE **`r sum(data$class=="target")`** TEST IMAGES.


- Performance:
```{r, cache=FALSE}
proc.time()
library(pryr)
mem_used()
```
