---
title: "Re-mining pipeline"
output: html_document
---

```{r, echo=F, warning=F}
screen <- "twoPlates"
channel <- "nucleus-dapi"
```

> **SCREEN:    `r screen`**

> **CHANNEL:   `r channel`**

------------------------------------------------------------------------------------------------------------------------------

### **Step 0: Set up the environment management**

------------------------------------------------------------------------------------------------------------------------------

- Screen settings.
```{r, echo=F}
if (screen == "mockScreen")
{
  INDEX_START_FEATURES <- 15
}
if (screen == "secretionPilot")
{
  INDEX_START_FEATURES <- 16
}
if (screen == "twoPlates")
{
  INDEX_START_FEATURES <- 14
}
```

- Load constants, paths and functions.
```{r, echo=F, message=F, warning=F}
source("includes/includeConstants.R")
source("includes/includePaths.R")
source("includes/includeFunctions.R")
library(ggplot2, quietly=T)
library(plotly, quietly=T)
```

------------------------------------------------------------------------------------------------------------------------------

### **Step 1: Feature extraction from images (WND-CHARM)**

------------------------------------------------------------------------------------------------------------------------------

- Data file must have the following column structure:
    - First part: metadata, where column order is not relevant, except for *class* that must be at the end.
    - Second part: undetermined number of features from WND-CHARM in any order.

```{r, echo=F}
dataPath <- file.path(pathTo_data, paste0(screen,"_db_",channel,".txt"))
data <- read.csv(dataPath, header=T, sep="\t", check.names=F)
```

- Standardize naming: TEST CLASS WITH NO PHENOTYPE.
```{r, echo=F}
# Elements with no phenotype (from test images) should have as label "NO_PH" and IC=0
if (length(data[data$specificCMPO==LABEL_NO_PHENOTYPE,]$IC) > 0)
{
    data[data$specificCMPO==LABEL_NO_PHENOTYPE,]$IC <- 0
    levels(data$tag) <- c(levels(data$tag), LABEL_NO_PHENOTYPE_TAG)
    data[data$specificCMPO==LABEL_NO_PHENOTYPE,]$tag <- LABEL_NO_PHENOTYPE_TAG
}
```

- Standardize labels: CONTROL CLASS.
```{r, echo=F}
# Those with empty gene are no-targeted genes
if (length(data[data$symbol=="",]$symbol) > 0)
{
    data[data$symbol=="",]$symbol <- LABEL_CONTROL_CLASS
}
```

- Detect the appropiate controls.
```{r, message=F, echo=F}
# Real controls
controlSiRNA <- c("103860","251283","scramble")
# There are controls in the dataset that are not controls
markedAsControlsInTheDataset <- data[data$class==LABEL_CONTROL_CLASS,]
# Those siRNA that are NOT in the real controls will be removed
wrongControls <- subset(markedAsControlsInTheDataset, !(siRNA %in% controlSiRNA))
# Remove wrong controls from the dataset
data <- utils.differenceOfDataFrames(data, wrongControls)
```

- Standardize data: CONTROL CLASS.
```{r, echo=F}
if (nrow(data[data$symbol=="",]) > 0)
{
    data[data$symbol==LABEL_CONTROL_CLASS,]$specificName <- LABEL_NO_PHENOTYPE
    data[data$symbol==LABEL_CONTROL_CLASS,]$specificCMPO <- LABEL_CONTROL_CLASS
    data[data$symbol==LABEL_CONTROL_CLASS,]$EnsemblID    <- LABEL_CONTROL_CLASS
    data[data$symbol==LABEL_CONTROL_CLASS,]$IC           <- 0
    data[data$symbol==LABEL_CONTROL_CLASS,]$tag          <- LABEL_CONTROL_TAG
    data[data$symbol==LABEL_CONTROL_CLASS,]$class        <- LABEL_CONTROL_CLASS
}
```

- Build an unique ID to identify images (*data$id*): **gene_siRNA_plate_replicate_well**.
```{r, echo=F}
data$id <- paste(data$symbol, data$siRNA, data$Plate, data$rep, data$Well, sep="_") 
data <- data[c(ncol(data), 1:(ncol(data)-1))]
INDEX_START_FEATURES <- INDEX_START_FEATURES + 1

colsMetadata <- colnames(data)[1:(INDEX_START_FEATURES-1)]
data.metadata <- data[c(colsMetadata[colsMetadata!="class"],colsMetadata[colsMetadata=="class"])]
data.features <- data[c(INDEX_START_FEATURES:ncol(data))]
data <- cbind(data.metadata, data.features)
rm(data.metadata, data.features)
```

- Number of images for each siRNA.
```{r, echo=F}
sirna_freq <- as.data.frame(table(data$siRNA))
colnames(sirna_freq) <- c("siRNA", "#images")
```

- Identify positive controls with id + "**********".
```{r, echo=F}
positiveControls <- subset(sirna_freq, `#images`>5)$siRNA
# Remove negative controls
positiveControls <- positiveControls[positiveControls!="103860"]
# Mark siRNA and gene names
data[data$siRNA %in% positiveControls,]$id <- paste0(data[data$siRNA %in% positiveControls,]$id,"**********")
```

- Save data.
```{r, echo=F}
standardizedDataPath <- file.path(pathTo_data, paste0(screen,"_standardized_db_",channel,".txt"))
write.table(data, file=standardizedDataPath, sep="\t", col.names=T, row.names=T, append=F, quote=F)
```

- Set colors for plots.
```{r, echo=F}
colorCodes <- unique(data[c("id","tag")])
colnames(colorCodes) <- c("label","tag")
```

- Structure pointers to adress data.
```{r, echo=F}
dataColNames <- colnames(data)
indexID      <- match("id", dataColNames)
indexClass   <- match("class", dataColNames) 
indexRangeFeature <- INDEX_START_FEATURES:ncol(data)
```

> THERE ARE **`r nrow(data)`** IMAGES.

```{r, echo=F}
table(data$class)
table(data$tag)
table(data$specificCMPO)
```

------------------------------------------------------------------------------------------------------------------------------

### **Step 2: Feature selection**

------------------------------------------------------------------------------------------------------------------------------

#### **2.1. Initial filtering of the features**

- Select only the WND-CHARM features (without metadata).
```{r, echo=F}
originalFeatures <- data[,indexRangeFeature]
```

- Remove constant columns (standard deviation=0).
```{r, echo=F}
features_noConstants <- stats.remove_constantColumns_fromDF(originalFeatures)
featureNamesToBeSelected <- names(features_noConstants)
```

- Update data and pointers.
```{r, echo=F}
metadata <- data[1:(INDEX_START_FEATURES-1)]
features <- data[featureNamesToBeSelected]

data <- data.frame(metadata, features, check.names=F)
indexRangeFeature <- c((ncol(metadata)+1):ncol(data))
```

#### **2.2. PCA for feature selection**

- Remove positive controls before PCA.
```{r, echo=F}
data <- data[!(data$siRNA %in% positiveControls),]
```

- Principal Component Analysis.
```{r, echo=F}
pca <- prcomp(data[indexRangeFeature], scale.=T, center=T)
```

- Eigenvalues of the PCs.
```{r, echo=F}
pcaEigenvalues <- pca$sdev^2
```

- Select the number of eigenvalues given by the elbow.
```{r, echo=F}
numberOfSelectedFeatures <- stats.get_elbow_ofDistribution(1:length(pcaEigenvalues), pcaEigenvalues)
```

- Plot eigenvalues.
```{r, fig.height=4, fig.width=8, message=F, echo=F}
plotEigenvalues <- as.data.frame(pcaEigenvalues)
plotEigenvalues$eigenvalue <- 1:length(pcaEigenvalues)
colnames(plotEigenvalues)[1] <- "value"

library(viridis)
g <- ggplot(plotEigenvalues, aes(x=eigenvalue, y=value))
g <- g + geom_point(shape=20, size=1)
g <- g + geom_vline(xintercept=numberOfSelectedFeatures, colour="red")
g <- g + ggtitle(paste0("VARIANCES [1, ",nrow(plotEigenvalues),"]"))
g <- g + scale_color_viridis(name="", discrete=T, end=0, begin=1)
g <- g + theme_bw()
ggplotly(g)

pcaEigenvalues <- pcaEigenvalues[1:numberOfSelectedFeatures]
```

| Stage                                           | # features                        |
| ----------------------------------------------- |:---------------------------------:|
| After WND-CHARM feature extraction              | `r ncol(originalFeatures)`        |
| After cleaning constant columns                 | `r ncol(features_noConstants)`    | 
| After PCA                                       | `r numberOfSelectedFeatures`      |
|                                                 |                                   | 

- Select `r numberOfSelectedFeatures` PCA features.
```{r, echo=F}
filteredFeatures <- as.data.frame(pca$x[,1:numberOfSelectedFeatures])
metadata <- data[1:(INDEX_START_FEATURES-1)]
```

- Update data and pointer with the new PCA features.
```{r, echo=F}
data <- data.frame(metadata, filteredFeatures, check.names=F)
indexRangeFeature <- INDEX_START_FEATURES:ncol(data)
```

- Save `r numberOfSelectedFeatures` PCA features.
```{r, echo=F}
PCAfeaturesFile <- file.path(pathTo_PCAs, paste0(screen,"_db_PCAs_",channel,".txt"))
write.table(data, file=PCAfeaturesFile, sep="\t", col.names=T, row.names=T, append=F, quote=F)
```

- Save PCA loadings.
```{r, echo=F}
loadingFile <- file.path(pathTo_PCAs, paste0(screen,"_PCAloadings_",channel,".txt"))
write.table(pca$rotation[,1:numberOfSelectedFeatures], file=loadingFile, sep="\t", col.names=T, row.names=T, append=F, quote=F)
```

------------------------------------------------------------------------------------------------------------------------------

### **Step 3: Quality control**

------------------------------------------------------------------------------------------------------------------------------

- Split data into control and test.
```{r, echo=F}
controlClass <- subset(data, data$class==LABEL_CONTROL_CLASS)
testClass    <- subset(data, data$class!=LABEL_CONTROL_CLASS)

rownames(controlClass) <- controlClass$id
rownames(testClass)    <- testClass$id

controlClassFeatures   <- controlClass[indexRangeFeature]
testClassFeatures      <- testClass[indexRangeFeature]
```

> THERE ARE **`r nrow(testClass)`** TEST IMAGES AND **`r nrow(controlClass)`** CONTROL IMAGES.

#### **3.1. Control Class definition (CC)**

- Distance between the control class images.
```{r}
covMatrix <- diag(pcaEigenvalues)
distControlImagesToCC <- mahalanobis(controlClassFeatures, colMeans(controlClassFeatures), covMatrix, tol=1e-20, inverted=F) # inverted=F when the cov matrix is not inverted outside
```

- Find a threshold distance with **25%** chance of missing a control.

```{r}
# Mahalanobis distance (d^2) is Chi-squared distributed.
percControlsRejected <- 25
alpha <- percControlsRejected/100
degreesOfFreedom <- numberOfSelectedFeatures
t <- qchisq(alpha, df=degreesOfFreedom, lower.tail=F)
```

> THRESHOLD DISTANCE = **`r t`**

- Select the control images outside of the control boundaries.
```{r, echo=F}
controlImages_farFromControls <- distControlImagesToCC[distControlImagesToCC>t]
```

| CONTROL IMAGES                                 | # images                               |
| --------------------------------------------|:-----------------------------------------:|
| Total number of control images              | `r nrow(controlClass)`                    |
| Number of control images far from the CC    | `r length(controlImages_farFromControls)` |
|                                             |                                           |
| **Number of remaining control images**      | **`r nrow(controlClass) -length(controlImages_farFromControls)`**     |
|                                             |                                           |

- Update the CC to have only those controls images close to the center of the CC.
```{r, echo=F}
controlClass <- controlClass[!(controlClass$id %in% names(controlImages_farFromControls)),]
controlClassFeatures <- controlClass[indexRangeFeature]
```

#### **3.2. Distance of Test Images to the Control Class**

- Mahalanobis distance of the test images to CC (note that this control class has been updated).
```{r, echo=F, message=F}
distTestImagesToCC <- mahalanobis(testClassFeatures, colMeans(controlClassFeatures), covMatrix, tol=1e-20, inverted=F)
```

- Reject test images with the Mahalanobis distance within the boundaries of the CC.
```{r, echo=F}
testImages_similarToControls <- distTestImagesToCC[distTestImagesToCC<=t]
```

| TEST IMAGES                                 | # images                                  |
| --------------------------------------------|:-----------------------------------------:|
| Total number of test images                 | `r nrow(testClassFeatures)`               |
| Number of test images similar to controls   | `r length(testImages_similarToControls)`  |
|                                             |                                           |
| **Number of remaining test images**         | **`r nrow(testClassFeatures)-length(testImages_similarToControls)`**                           |
|                                             |                                           |

#### **3.3. Histograms of control and test images**

```{r, fig.height=3, fig.width=9, message=F, echo=F}
library(viridis)
g <- qplot(distControlImagesToCC, binwidth=1, xlab="Distance of the control images to the center of the Control Class")
g <- g + geom_vline(xintercept=t, colour="green")
g <- g + scale_color_viridis(name="", discrete=T, end=0, begin=1)
g <- g + theme_bw()
ggplotly(g)

g <- qplot(distTestImagesToCC, binwidth=10, xlab="Distance of the test images to the center of the Control Class")
g <- g + geom_vline(xintercept=t, colour="green")
g <- g + scale_color_viridis(name="", discrete=T, end=0, begin=1)
g <- g + theme_bw()
ggplotly(g)
```

#### **3.4. Data reduction**

- Reject ALL CONTROLS (**-`r nrow(controlClassFeatures)`**): 

Starting with **`r nrow(data)`** images (TI + CC)...
```{r}
data <- data[data$class!=LABEL_CONTROL_CLASS,]
```
... we get **`r nrow(data)`** (TI + CC).

```{r, echo=F}
table(data$tag)
```

- Reject TEST IMAGES close to controls (**-`r length(testImages_similarToControls)`**):
Starting with **`r nrow(data)`** images (TI + CC)...
```{r}
testImagesToReject <- as.factor(names(testImages_similarToControls))
data <- data[!data$id %in% testImagesToReject,]
```
... we get **`r nrow(data)`**.

```{r, echo=F}
table(data$tag)
```

- Reproducibility:
Starting with **`r nrow(data)`** images (TI + CC)...
```{r}
X <- 2
data <- images.remove_siRNAWithLessThanXImagesLeft(data, X)
```
... we get **`r nrow(data)`**.

```{r, echo=F}
table(data$tag)
utils.sort_vector(table(data$specificName))
```

- Save data.
```{r, echo=F}
fileData <- file.path(pathTo_QC, paste0(screen,"_db_qualityControlled.txt"))
write.table(data, file=fileData, sep="\t", col.names=F, row.names=T, append=F, quote=F)
```

#### **3.5. Plot the remaining images on the first two PCs.**

```{r, fig.height=10, fig.width=10, echo=F}
library(viridis)
g <- ggplot(data, aes(x=PC1, y=PC2, color=tag, text=id))
g <- g + geom_point(shape=20, size=1.5)
g <- g + scale_color_viridis(name="", discrete=T, end=0, begin=1)
g <- g + theme_bw()
ggplotly(g)
```

### **Step 4: Similarity between images**

- Cosine similarity between PCA features of the test images.
```{r, echo=F}
features <- data[c(INDEX_START_FEATURES:ncol(data))]
```

- Similarity matrix.
```{r, echo=F, warning=F, message=F}
simMatrix <- measures.get_cosineSimilarityMatrix(features)
rownames(simMatrix) <- colnames(simMatrix) <- data$id
```

- Reorder matrix by gene to have them together in the heat map.
```{r, echo=F}
newOrder  <- rownames(simMatrix)[order(rownames(simMatrix))]
simMatrix <- simMatrix[newOrder, newOrder]
```

- Save heat map of the similarity matrix between images.
```{r, echo=F, warning=F, message=F}
simList <- utils.convert_MatrixToListOfPairs(simMatrix)

library(viridis)
gg <- ggplot(simList, aes(x=X1, y=X2, fill=value))
gg <- gg + geom_tile(color="white", size=0.15)
gg <- gg + scale_fill_viridis(name="similarity")
gg <- gg + coord_equal()
gg <- gg + labs(x=NULL, y=NULL)
gg <- gg + theme(axis.text.x = element_text(angle=90, hjust=1))
gg <- gg + theme(axis.ticks=element_blank())
gg <- gg + theme(axis.text=element_text(size=7))
gg <- gg + theme(legend.title=element_text(size=12))
gg <- gg + theme(legend.text=element_text(size=10))
ggplotly(gg, width=1000, height=1000)

ggsave(gg, width=20, height=20, filename=file.path(pathTo_plots,"cosineSimilarityBetweenImages.pdf"))
```

- Performance:
```{r, cache=FALSE, echo=F}
proc.time()
library(pryr)
mem_used()
```
